The Effects of the NBA COVID Bubble on the NBA Playoffs

### 2022-02-02 Manuscript/NBABubbleAnalysis.Rmd
Update NBABubbleAnalysis.Rmd
Generally proofreading, grammar and flow check

---
title: "The Effects of the NBA COVID Bubble on the NBA Playoffs"
subtitle: "A Case Study for Home-Court Advantage"
author: |
  | Michael Price and Jun Yan
  | michael.price@uconn.edu
  | Department of Statistics, University of Connecticut
date: "`r format(Sys.time(), '%d, %B, %Y')`"
documentclass: article
papersize: letter
fontsize: 11pt
bibliography: citations.bib
biblio-style: asa
output: 
  bookdown::pdf_document2:
    keep_tex: yes
header-includes:
   - \usepackage[margin=1in]{geometry}
   - \usepackage {hyperref}
   - \hypersetup {colorlinks = true, linkcolor = blue, citecolor=blue, urlcolor = blue}
toc: False
---
Abstract:
The 2020 NBA playoffs were played inside of a bubble in Disney World because of
the COVID-19 pandemic. This meant that there were no fans in attendance,
games played on neutral courts and no traveling for teams, which in theory
removes home-court advantage from the games. This setting has attracted much
discussion as analysts and fans debated the possible effects it may have on the
outcome of games. Home-court advantage has historically played an influential
role in NBA playoff series outcomes. The 2020 playoff provided a unique
opportunity to study the effects of the bubble and home-court
advantage by comparing the 2020 season with the seasons in the past.
While many factors contribute to the outcome of games, points scored
is the deciding factor of who wins games, so scoring is the primary focus
of this study. The specific measures of interest are team scoring totals and
team shooting percentage on two-pointers, three-pointers, and free throws.
Comparing these measures for home teams and away teams in 2020 vs. 2017-2019
shows that the 2020 playoffs favored away teams more than usual, particularly
with two point shooting and total scoring.

@@

# Introduction {#sec:intro}
Home-court advantage is often discussed in sports circles as a contributing
factor to the outcome of games. It is well-known that the home team typically
benefits from some competitive edge from playing at their home-court, resulting
in a better chance of winning. Thus, the NBA playing the 2020 playoffs in a
bubble due to the COVID-19 pandemic
brought a great deal of concern for fans, teams, journalists, and others.
Among the biggest is the effect from the crowd, a home crowd has the ability
to energize home teams, help control momentum in favor of the home team and 
create a chaotic and stressful environment for away teams to play in. There 
is no way replicate these effects in the bubble conditions. Also, there are 
geographic factors that are lost like altitude and time-zone effects. There's 
evidence teams across all sports in high altitude regions can rely on away
teams coming in and struggling with lower oxygen levels at a higher altitude, 
which the home teams are accustomed to playing with \citet{Lopez}. Examples,
of high altitude teams in the NBA would be the Utah Jazz and Denver Nuggets, 
both of whom participated in the 2020 NBA bubble. Also, referees are known to
be subject to favoring home teams in large part due pressure from fans, which
is no longer in play in the bubble. Refs may still have biases, but there's
nothing to sway that bias towards the home team like normal. Any positive 
effects of the home team playing in the arena they're most comfortable in 
and living in the comfort of their own home are also lost. Travel is also 
normally discussed as a factor hurting away teams, but for the context of
the playoffs in the bubble losing this may not be significant. In the
playoffs, many times (aside from the first game of the series when the 
away team travels) both teams often travel at the same time since they're 
both going back and forth between the two cities to complete the series.

@@

Our study is quite different from earlier NBA home-court advantage studies.
By using the neutral site games of 2020 we will get to compare home and away
performance to a control. Typically, studies just compare home vs away
performance. These studies do not separate the effects of home-court advantage
into the specific effect on the home team and the specific effect on the away
team. They show that home teams outperform away teams, but not if this is a
result of home teams overperforming or away teams underperforming because of
home-court advantage. Some of these studies are reviewed in greater detail in
Section~\@ref(sec:litrev).

@@

# Literature Review {#sec:litrev}
There is a voluminous literature on the effects of home-court advantage.
Many NBA home-court advantage studies analyze the effects by studying shooting
percentages. @Kotecki reported significant home-court advantage using performance-based
statistics, specifically field goal percentage, free throw percentage, and
points by comparing home performance vs. away performance in games.
All of which he showed significantly indicates that
home-court helps teams play better. @Cao studied the effects
of pressure on performance in the NBA. Using free throws as their measure of
interest, they tested whether home
fans could distract and put pressure on opposing players to make free
throws. However, they did not find significant evidence
evidence that home status has a substantial impact on missing from the free throw
line. @Harris used two point shots, three-point shots and free throws as measures
of interest to study home-court advantage. Two point shots were found to be
the strongest predictor of home-court advantage. They suggested
that home teams should try to shoot more two point shots and force
their opponent to take more two point shot attempts. This minimum variance strategy will 
help home teams have a greater control of the game play and help maximize the benefits 
home court advantange.

Some studies focus less on shooting and more on scoring differences and other metrics.
For example, @Greer focused on the influence of spectator booing on
home-court advantage in basketball. The three methods of performance used
in this study were scoring, violations, and turnovers. This study was conducted
using the men’s basketball programs at two large universities. The study finds that
social support, like booing, is an important contributor to home-court advantage.
Greer explains, whether the influence is greater on visiting team performance or
referee calls is less clear. However, the data does seem to lean slightly in
favor of affecting visiting team performance.  @Harville studied the
effect of home-court advantage using the 1991-1992 college basketball season.
Unlike the NBA, it is not uncommon to have a few games played at neutral sites
during the college basketball season. This allowed them to construct two samples,
one of home teams and one of neutral teams. They formulate their study in a
regression predicting the expected difference in score for home teams.
They set up their study to find if the home teams won games by more points when
they had home-court advantage vs. when playing at neutral court. This study
concluded with evidence supporting home-court advantage.

There are also surveys on the factors contributing to home-court advantage.
@Carron1992 gave four main game location factors for home and away
teams, namely, the crowd factor, which is the impact of fans cheering; learning
factors, which is an advantage from home teams from playing at
a familiar venue; travel factors, the idea that away teams may face
fatigue and jet lag from traveling; and, rule factors, which says that home teams
may benefit from some advantages in rules and officiating. They acknowledge that
these factors would all be removed if games were played at a neutral site even
if one team was designated as “home team”. This study was reviewed a decade later
by @Carron2005. The 2005 review goes over the new findings from studies
about the significance of these four game location factors. Since 1992 they found
that results on these four factors are mixed. However, there is some evidence
supporting crowd and travel factors impact games across all major sports. There is less
evidence suggesting learning and rule factors impact across the various collegiate 
and professional sports, regardless, the NBA is not a league which has rules that
may favor the home team like batting last in the MLB. One interesting finding cited
by @Carron2005 is that in the absence of crowds result in overall performance increases.
Lastly, as discussed in the introduction @Lopez find evidence that geographic
factors like altitude may influence and strengthen home court advantage for teams in
high altitude regions.

@@

# Data {#sec:data}
Data were collected from the official NBA website. The main variables of interest
are whether or not the home team won, scoring totals for home and away teams, and
shooting percentages for home and away teams on two-pointers, three-pointers, and
free throws. These variables were very popular and frequently used in the related
literature discussed earlier. While many other measures could be used for examining
the outcome of the game and team performance, scoring seemed to be the most
important. The winner of a game is determined by who scores more points. The three 
types of shots are anatural discussion point in basketball, as they're all
important and directly impact scoring totals. Free throws can often determine the winner of close
games, as losing teams are forced to foul the winning team to
stop the clock and hope for some missed free throws. The three
point shot has grown very prominent in basketball as the entire NBA has increased
their volume of three point shot attempts in response to the
recent success of the Golden State Warriors and Stephen Curry. Due to this fact, 
there is a growing consensus the three point shot is most important in basketball. 
However, two point shots are not to be overlooked and may
actually be most important, especially for home teams, according
to \citet{Harris}. In fact, 2020 NBA champion Los Angeles Lakers actually led league
in two point percentage throughout the playoffs and were bottom five in three point percentage.

@@

Home-court advantage is the basic idea that the home team is more likely to win.
So laying a foundation of typical home-court advantage is crucial. Before
focusing on the 2017 to 2020 playoffs we can take a quick look at home team win
percentages since 2010. Notice in Figure \@ref(fig:Fig1), the 10 years before 2020, the home team
winning percentage ranged from around 0.56 to 0.7 and never dipped below 0.5. The
2020 bubble broke this historic pattern dipping down below 0.5. Foreshadowing,
the confirmation of the expectation that the effect of home-court advantage was removed
in the 2020 playoffs.

@@

Moving on to the main focus of the study, comparing 2020 to 2017-2019.
Figure \@ref(fig:Fig2) shows the histograms of the home (green) and
away scoring (red) for 2020 vs. 2017-2019. All histograms are
fairly bell shaped, which is important for statistical tests
designed for normally distribute data. There appears to be little
difference between the 2020 and 2017-2019 for home scoring, but for
away scoring, a more pronounced shift to the right in 2020 is observed
compared with that in 2017-2019.

@@

Our second target of inference is shooting percentage for home and away teams.
Figure \@ref(fig:Fig3) shows home shooting for two-pointers, three-pointers and
free throws for 2020 (top) vs. 2017-19 (bottom). The histograms appear to be fairly
similarly distributed between 2020 and 2017-19. Likewise, Figure \@ref(fig:Fig4),
shows the same percentages except for away teams. It appears that the
two-point shooting percentage for away teams has a small shift to the
right in 2020 relative to that in 2017-2019.

@@

# Methods {#sec:methods}
The 2020 bubble provides a new and exciting opportunity to study home-court
advantage for the NBA. Unlike college basketball, aside from a few
exhibition/preseason games, the NBA always has a home and away team. So, for
the first time in NBA history the bubble allows NBA home and away performance
to be compared vs a control/neutral field. The NBA bubble, as a neutral court,
removed all 4 possible game location factors impacting home-court advantage
hypothesized by @Carron1992. The NBA bubble featured 8 seeding games then a
standard playoff format. The focus of this study was the play during the playoff
games since it followed the standard playoff format and can easily be compared
back to other playoffs. For this study, the 2020 playoffs were compared against
the three previous playoffs collectively. To control
for the changing play style of the NBA, we limit the study to 2020 vs 2017-2019
for the faster pace play and more common use of the three-point shot
in modern basketball. If we used data from say 10 years ago, or earlier, observed
differences may not be from effects of the NBA bubble, but rather from the effects
of drastic changes in the style of play between the seasons. However, basketball
evolves slow enough that we can reasonably assume 2017-2019 are at least very
close in pace and playing style to 2020.

Comparisons between 2020 and 2017-19 home and away teams were made on home team
winning percentage, total team scoring and two-point, three-point and free throw
shooting. Comparing the differences in these metrics for home and away teams in
2020 vs previous years will provide valuable insights to the understanding of
home-court advantage. We can see how going on the road may negatively impact
away performance and how playing at home may positively impact home performance.
If there are differences in scoring for home or away teams the differences can
be used to show how home-court advantage affects overall performance of home
and away teams. While testing for differences in shooting will provide added
context for how home-court advantage specifically affects performance.
Shooting percentages are not the only possible metrics affected by home-court
advantage, but they are the most obvious and certainly an important one.

@@

We also conducted nonparametric tests that are distribution free to
confirm the results from the $z$-test. For question~1, we used
Fisher's exact test for a contigency table which summarizes the wins
and losses of the home team in the 83 games in 2020 and the 243 games
in 2017-2019. For all other eight questions, the data are the scores
or shooting percentages from the 84 games in 2020 and the 243 games in
2017-2019. We used Wilcoxan's rank-sum test.

@@

Starting from the top of the 
Table \@ref(tab:table) summarizes p-values of the nine hypotheses for both
$z$-tests and Wilcoxon tests. The p-values are all fairly similar for both
tests giving strong confidence in conclusions drawn. Also reported are 
the point estimates of the two samples in each comparison.

@@

# Discussion {#sec:disc}
Generally it seemed that away teams fared better in the 2020 NBA playoff bubble
than previous years on the road. Starting from the dip in home winning percentage
to below 0.482 it is clear that something was different. Although the difference was not
significant after a Bonferroni correction it is still informative to consider and
understand that home teams seemed to struggle to win compared to normal conditions.
Compared to @Kotecki who finds home teams consistently have a significantly
better record than away teams boasting about a 60.5% win percentage in his sample,
the 48.2% home winning percentage of 2020 home teams is quite a shift.
In this study, home teams did not benefit from the usual advantages provided by
being the home team.

@@

At least some of that improvement from away teams came from significantly higher
two-point efficiency. This corresponds with the conclusion from \citet{Harris},
where they found home teams are best suited to capitalize on advantages from two-point
shots. Normally, by shooting more two-pointers themselves and forcing away teams to
shoot more two-pointers the home team benefits from increasing effects of
home-court advantage. However, with away teams significantly improving two-point
shooting in the bubble this strategy was no longer viable and home-court
advantage disappeared.

To get a bit more detailed, we saw an estimated 3\% increase in away two point
shooting percentage. Likewise, NBA teams shot
an average of about 53 two point shots per game, in our data, leading to a three 
point increase in total scoring. Even this
small swing in scoring for away teams can make a big difference in outcome of
games. Around 1/3 of NBA games end by a decision of six points or less, those
are the games where the three points matter most. Suppose there are 30 seconds
left and the away team is losing. If the away team has the ball, only down three, 
they can take their time finding a good shot which can
still either be a two or three, given the score and remaining time. In the context 
of the NBA bubble, this crucial possession also has the benefit of the away team
getting to work in silence without the jeers from enemy fans. However, in the
same situation while being down six the offense must rush to get a quick shot
off, which probably has to be a three pointer. This is a much more desperate
circumstance, less likely to have a positive outcome and even in the positive
outcome case you're still going to be down three and likely have to foul the
opposing team after you score. 
Alternatively, if the away team is playing defense with 30 seconds left, only down 
three, you can play regular defense without fouling, get a stop, then draw up a play 
to score a three pointer with the remaining six seconds. However, in the same scenario 
down six, you have no choice, but to foul. The free throws will likely put you down
seven or eight points, but the alternative is let the opponent run the 24 second shot 
clock down wasting time. Also, the away team may even be in a situation where they're
leading by the three additional points, which for the same logic as explained above 
puts them in far greater control of the game. Clearly, the three point swing can 
drastically change probability of winning for a team down the stretch of a game. 
Since, home teams picked up an additional two points per game on average vs. five 
for away teams, the three additional points from two point shooting appear to be the 
driving factor of away teams closing the gap with home teams.

@@

Future studies may want to use the 2020 NBA bubble and compare vs previous years
using other performance measures. For example, turnovers, steals, assist, rebounds,
and many more game statistics. There are plenty of other possibilities besides
just shooting efficiency to pick through looking for more possible sources of
added points for away teams. This will further help explain what is lost in the
performance of away teams when they travel to opposing arenas. This study is only
the beginning of possibilities for studies using the 2020 NBA bubble as a case study
for home-court advantage. Although the study is limited by a one time
sample, it seems unlikely that these conditions will ever be
repeated. It may not be possible to have a follow-up study using the
same measures with a different sample. Otherwise, that type of study
could help strengthen the conclusion in this paper. Also, you could
answer more questions with a larger sample size. For example,  testing
strength of home court advantage relative to specific teams, like the
Denver Nuggets and Utah Jazz who we know from other literature may
have a stronger home court advantage. The sample size of home games
for the individual teams in the bubble was far too small to try and
address this question, the Jazz and Nuggets only had 3 and 9 home
games, respectively.

>>>>

---
title: "The Effects of the NBA COVID Bubble on the NBA Playoffs"
subtitle: "A Case Study for Home-Court Advantage"
author: |
  | Michael Price and Jun Yan
  | michael.price@uconn.edu
  | Department of Statistics, University of Connecticut
date: "`r format(Sys.time(), '%d, %B, %Y')`"
documentclass: article
papersize: letter
fontsize: 11pt
bibliography: citations.bib
biblio-style: asa
output: 
  bookdown::pdf_document2:
    keep_tex: yes
header-includes:
   - \usepackage[margin=1in]{geometry}
   - \usepackage {hyperref}
   - \hypersetup {colorlinks = true, linkcolor = blue, citecolor=blue, urlcolor = blue}
toc: False
---
Abstract:
The 2020 NBA playoffs were played inside of a bubble at Disney World because of
the COVID-19 pandemic. This meant that there were no fans in attendance,
games were played on neutral courts and no traveling for teams. In theory,
this should remove home-court advantage from the games. This setting genereated
discussion and concerns, as analysts and fans debated the possible effects it may have on the
outcome of games. Home-court advantage has historically played an influential
role in NBA playoff series outcomes. The 2020 playoffs provided a unique
opportunity to study the effects of the bubble and home-court
advantage by comparing the 2020 season with the seasons in the past.
While many factors contribute to the outcome of games, points scored
is the deciding factor of who wins. Thus, scoring is the primary focus
of this study. The specific measures of interest are team scoring totals and
team shooting percentage on two-pointers, three-pointers, and free throws.
Comparing these measures for home teams and away teams in 2020 vs. 2017-2019
shows that the 2020 playoffs favored away teams more than usual, particularly
with two point shooting and total scoring.

@@

# Introduction {#sec:intro}
Home-court advantage is often discussed in sports circles as a contributing
factor to the outcome of games. It is well-known that the home team typically
benefits from some competitive edge while playing at their home-court, resulting
in a better chance of winning. Thus, the NBA playing the 2020 playoffs in a
bubble due to the COVID-19 pandemic
brought a great deal of concern for fans, teams, journalists, and others. The bubble environment
would not be able to replicate many of the factors that normal home-court advantage relies on.
Among the biggest is the effect from the crowd. The home crowd has the ability
to energize home teams, help control momentum in favor of the home team and 
create a chaotic and stressful environment for away teams to play in. There 
is no way replicate these effects in the bubble conditions. Also, there are 
geographic factors that are lost, like altitude and time-zone effects. There's 
evidence across all sports that teams in high altitude regions can rely on away
teams coming in and struggling with lower oxygen levels at a higher altitude, 
which the home teams are accustomed to playing with \citet{Lopez}. Examples,
of high altitude teams in the NBA would be the Utah Jazz and Denver Nuggets, 
both of whom participated in the 2020 NBA bubble. Also, referees are known to
be subject to favoring home teams. This is due in large part to pressure from fans, which
is no longer in play in the bubble. Refs may still have biases, but there's
nothing to sway that bias towards the home team like normal. Any positive 
effects of the home team playing in the arena they're most comfortable in 
and living in the comfort of their own home are also lost. Travel is also 
normally discussed as a factor hurting away teams, but for the context of
the playoffs in the bubble, losing this may not be significant. In the
playoffs, many times (aside from the first game of the series when the 
away team travels) both teams often travel at the same time since they're 
both going back and forth between the two cities to complete the series.

@@

Our study is quite different from earlier NBA home-court advantage studies.
By using the neutral site games of 2020, we will get to compare home and away
performance to a control. Typically, studies just compare home vs away
performance. These studies do not separate the effects of home-court advantage
into the specific effect on the home team and the specific effect on the away
team. They show that home teams outperform away teams, but not if this is a
result of home teams overperforming or away teams underperforming because of
home-court advantage. Some of these studies are reviewed in greater detail in
Section~\@ref(sec:litrev).

@@

# Literature Review {#sec:litrev}
There is voluminous literature on the effects of home-court advantage.
Many NBA home-court advantage studies analyze the effects by studying shooting
percentages. @Kotecki reported significant evidence of home-court advantage by 
comparing field goal percentage, free throw percentage
and points scored in home vs. away teams. He found all of these measures showed evidence that
home-court advantage helps home teams play better. @Cao studied the effects
of pressure on performance in the NBA. Using free throws as their measure of
interest, they tested whether home
fans could distract and put pressure on opposing players to make free
throws. However, they did not find significant evidence
evidence that home status has a substantial impact on missing from the free throw
line. @Harris used two-point shots, three-point shots and free throws as measures
of interest to study home-court advantage. Two point shots were found to be
the strongest predictor of home-court advantage. They suggested
that home teams should try to shoot more two point shots and force
their opponent to take more two point shot attempts. This will 
help home teams have a greater control of the game play and help maximize the benefits 
home court advantange.

Some studies focus less on shooting and more on other metrics.
For example, @Greer focused on the influence of spectator booing on
home-court advantage in basketball. The three methods of performance used
in this study were scoring, violations, and turnovers. This study was conducted
using the men’s basketball programs at two large universities. The study finds that
social support, like booing, is an important contributor to home-court advantage.
Greer explains, whether the influence is greater on visiting team performance or
referee calls is less clear. However, the data does seem to lean slightly in
favor of affecting visiting team performance.  @Harville studied the
effect of home-court advantage using the 1991-1992 college basketball season.
Unlike the NBA, it is not uncommon to have a few games played at neutral sites
during the college basketball season. This allowed them to construct two samples,
one of home teams and one of neutral teams. They formulate their study in a
regression predicting the expected difference in score for home teams.
They set up their study to find if the home teams won games by more points when
they had home-court advantage vs. when playing at neutral court. This study
concluded with evidence supporting home-court advantage.

There are also surveys on the factors contributing to home-court advantage.
@Carron1992 gave four main game location factors for home and away
teams, namely, the crowd factor, which is the impact of fans cheering; learning
factors, which is an advantage from home teams from playing at
a familiar venue; travel factors, the idea that away teams may face
fatigue and jet lag from traveling; and, rule factors, which says that home teams
may benefit from some advantages in rules and officiating. They acknowledge that
these factors would all be removed if games were played at a neutral site even
if one team was designated as “home team”. This study was reviewed a decade later
by @Carron2005. The 2005 review goes over the new findings from studies
about the significance of these four game location factors. Since 1992 they found
that results on these four factors are mixed. However, there is some evidence
supporting crowd and travel factors impact games across all major sports. There is less
evidence suggesting learning and rule factors impact across the various collegiate 
and professional sports. The NBA is not a league which has rules that
may favor the home team, like batting last in the MLB, but these rule factors 
also account for referee bias which may impact the NBA. One interesting finding cited
by @Carron2005 is that in the absence of crowds result in overall performance increases.
Lastly, as discussed in the introduction @Lopez find evidence that geographic
factors like altitude may influence and strengthen home court advantage for teams in
high altitude regions.

@@

# Data {#sec:data}
Data were collected from the official NBA website. The main variables of interest
are whether or not the home team won, scoring totals for home and away teams, and
shooting percentages for home and away teams on two-pointers, three-pointers, and
free throws. These variables were very popular and frequently used in the related
literature discussed earlier. While many other measures could be used for examining
the outcome of the game and team performance, scoring seemed to be the most
important because the winner of a game is determined by who scores more points. Furthermore, the three 
types of shots are a natural discussion point in basketball, as they're all
important and directly impact scoring totals. Free throws can often determine the winner of close
games, as losing teams are forced to foul the winning team to
stop the clock and hope for some missed free throws. The three
point shot has grown very prominent in basketball. The entire NBA has increased
their volume of three point shot attempts in response to the
recent success of the Golden State Warriors and Stephen Curry.  Due to this
fact, there is a growing consensus the three point shot is crucial because
of the efficient scoring and floor spacing it provides offenses.  
However, two point shots are not to be overlooked and may
actually be most important, especially for home teams, according
to \citet{Harris}. In fact, 2020 NBA champion Los Angeles Lakers actually led league
in two-point shooting percentage throughout the playoffs and were bottom 
five in three-point shooting percentage.

@@

Home-court advantage is the basic idea that the home team is more likely to win given 
they benefit from positive effects of a few factors we discussed earlier.
So laying a foundation of typical home-court advantage is crucial. Before
focusing on the 2017 to 2020 playoffs we can take a quick look at home team win
percentages since 2010. Notice in Figure \@ref(fig:Fig1), the 10 years before 2020, the home team
winning percentage ranged from around 0.56 to 0.7 and never dipped below 0.5. The
2020 bubble broke this historic pattern dipping down below 0.5. Foreshadowing,
the confirmation of the expectation that the effect of home-court advantage was removed
in the 2020 playoffs.

@@

Moving on to the main focus of the study, comparing 2020 to 2017-2019.
Figure \@ref(fig:Fig2) shows the histograms of the home (green) and
away scoring (red) for 2020 vs. 2017-2019. All histograms are
fairly bell shaped, which is important for statistical tests
designed for normally distributed data. There appears to be little
difference between the 2020 and 2017-2019 for home scoring. For
away scoring, a more pronounced shift to the right in 2020 is observed
compared with that in 2017-2019.

@@

Our second target of inference is shooting percentage for home and away teams.
Figure \@ref(fig:Fig3) shows home shooting for two-pointers, three-pointers and
free throws for 2020 (top) vs. 2017-19 (bottom). The histograms appear to be fairly
similarly distributed between 2020 and 2017-19. Likewise, Figure \@ref(fig:Fig4),
shows the same percentages except for away teams. It appears that the
two-point shooting percentage for away teams has a small shift to the
right in 2020 relative 2017-2019.

@@

# Methods {#sec:methods}
The 2020 bubble provides a new and exciting opportunity to study home-court
advantage for the NBA. Unlike college basketball, aside from a few
exhibition/preseason games, the NBA always has a home and away team. So, for
the first time in NBA history, the bubble allows NBA home and away performance
to be compared against a control/neutral field. The NBA bubble
removed many, if not all, of the factors impacting home-court advantage. The NBA bubble featured 8 seeding games 
for each team, then a standard playoff format. The focus of this study was on the play during the playoff
games since it followed the standard playoff format and could be compared
back to other playoffs. For this study, the 2020 playoffs were compared against
the three previous playoffs collectively. To control
for the changing play style of the NBA, we limit the study to 2020 vs 2017-2019
due to the faster pace play and more common use of the three-point shot
in modern basketball. If we used data from say 10 years ago, or earlier, observed
differences may not be from effects of the NBA bubble, but rather from the effects
of drastic changes in the style of play between the seasons. However, basketball
evolves slow enough that we can reasonably assume 2017-2019 are at least very
close in pace and playing style to 2020.

Comparisons between 2020 and 2017-19 home and away teams were made on home team
winning percentage, total team scoring and two-point, three-point and free throw
shooting. Comparing the differences in these metrics for home and away teams in
2020 vs previous years will provide valuable insights to the understanding of
home-court advantage. We can see how going on the road may negatively impact
away performance and how playing at home may positively impact home performance.
If there are differences in scoring for home or away teams, the differences can
be used to show how home-court advantage affects overall performance of home
and away teams. While testing for differences in shooting will provide added
context for how home-court advantage specifically affects performance.
Shooting percentages are not the only possible metrics affected by home-court
advantage, but they are the most obvious and likely most important one.

@@

We also conducted nonparametric tests that are distribution free to
confirm the results from the $z$-test. For question 1, we used
Fisher's exact test for a contingency table which summarizes the wins
and losses of the home team in the 83 games in 2020 and the 243 games
in 2017-2019. For all other eight questions, the data are the scores
or shooting percentages from the 83 games in 2020 and the 243 games in
2017-2019. We used Wilcoxan's rank-sum test.

@@

Starting from the top of the 
Table \@ref(tab:table) summarizes p-values of the nine hypotheses for both
$z$-tests and Wilcoxon tests. The p-values are all fairly similar for both
tests giving strong confidence in conclusions drawn. Also, reported are 
the point estimates of the two samples in each comparison.

@@

# Discussion {#sec:disc}
Generally, it seemed that away teams fared better in the 2020 NBA playoff bubble
than previous years on the road. Starting from the dip in home winning percentage
to below 0.482, it is clear that something was different. Although, the difference was not
significant after a Bonferroni correction it is still informative to consider and
understand that home teams seemed to struggle to win compared to normal conditions.
Compared to @Kotecki, who finds home teams consistently have a significantly
better record than away teams boasting about a 60.5% win percentage, in his sample,
the 48.2% home winning percentage of 2020 home teams is quite a shift.
In this study, home teams did not appear to benefit from the usual advantages provided by
being the home team.

@@

At least some of that improvement from away teams came from significantly higher
two-point efficiency. This corresponds with the conclusion from \citet{Harris},
where they found home teams are best suited to capitalize on advantages from two-point
shots. Normally, by shooting more two-pointers themselves and forcing away teams to
shoot more two-pointers, the home team benefits from increasing effects of
home-court advantage. However, with away teams significantly improving two-point
shooting in the bubble, this strategy was no longer viable and home-court
advantage disappeared.

To get a bit more detailed, we saw an estimated 3\% increase in away two-point
shooting percentage. Likewise, NBA teams shot
an average of about 53 two point shots per game, in our data, leading to a three 
point increase in total scoring. Even this
small swing in scoring for away teams can make a big difference in outcome of
games. Around 1/3 of NBA games end by a decision of six points or less, those
are the games where the three points matter most. Suppose there are 30 seconds
left and the away team is losing. If the away team has the ball, only down three, 
they can take their time finding a good shot which can
still either be a two or three given the score and remaining time. In the context 
of the NBA bubble, this crucial possession also has the benefit of the away team
getting to work in silence without the jeers from enemy fans. However, in the
same situation while being down six the offense must rush to get a quick shot
off, which probably has to be a three pointer. This three point swing creates a much more desperate
circumstance which is less likely to have a positive outcome. Even in the positive
outcome case you're still going to be down three and likely have to foul the
opposing team after you score. 
Alternatively, if the away team is playing defense with 30 seconds left, only down 
three, you can play regular defense without fouling, get a stop, then draw up a play 
to score a three pointer with the remaining six seconds. However, in the same scenario 
down six, you have no choice, but to foul. The free throws will likely put you down
seven or eight points, but the alternative is let the opponent run the 24 second shot 
clock down wasting time. Similarly, the away team may even be in a situation where they're
leading by the three additional points, which for the same logic as explained above 
puts them in far greater control of the game. Clearly, the three point swing can 
drastically change probability of winning for a team down the stretch of a game. 
Since, home teams picked up an additional two points per game on average vs. five 
for away teams, the three additional points from two point shooting appear to be the 
driving factor of away teams closing the gap with home teams.

There are a few possibilities that might create this increase in two-point shooting percentage. 
One could be the possible removal of officiating bias in favor of the home team during the
bubble. Away teams in the bubble may have benefited from more fouls called on drives to
the basket. On drives to the basket there can often be a lot of contact with no clear
foul, these calls are then up to the ref's discretion. Typically, refs may be more
reluctant to blow the whistle against the home team, in front of their crowd, on this 
type of play. The NBA only records an attempted shot when a foul is called on the shot
if the shot is made. So, when away teams are fouled, but it is not called because it's 
a often at the ref's discretion to make the call, they'll likely miss and be credited the 
miss. However, in the bubble if fouls are called more fairly, the missed shots from uncalled
fouls are removed and two-point shooting would increase. This wouldn't affect three-point 
shooting because those fouls are more clear and less up to ref's discretion. Another possibility
, without the crowd noise inhibiting their offense, away teams were able to more easily
run their offensive sets that generate easy two-point looks at the rim. Also, away teams
being generally more confident without opposing fans present, may have been more inclined
to attack the basket and get an easy look close to the rim. It's hard to say for sure what 
cause the increase in away two-point shooting, but these are all possibilities. 
It may also be a combination of all of these.

@@

Future studies may want to use the 2020 NBA bubble and compare vs previous years
using other performance measures. For example, turnovers, steals, assist, rebounds,
and many more game statistics. There are plenty of other possibilities besides
just shooting efficiency to pick through looking for more possible sources of
added points for away teams. This will further help explain what is lost in the
performance of away teams when they travel to opposing arenas. This study is only
the beginning of possibilities for studies using the 2020 NBA bubble as a case study
for home-court advantage. Although, the study is limited by a one time
sample, it seems unlikely that these conditions will ever be
repeated. It may not be possible to have a follow-up study using the
same measures with a different sample. Otherwise, that type of study
could help strengthen the conclusion in this paper. Also, you could
answer more questions with a larger sample size. For example,  testing
strength of home court advantage relative to specific teams, like the
Denver Nuggets and Utah Jazz who we know from other literature may
have a stronger home court advantage. The sample size of home games
for the individual teams in the bubble was far too small to try and
address this question, the Jazz and Nuggets only had 3 and 9 home
games, respectively.
